\section{Related work}
\label{sec:related}

A large body of work focuses on using shared memory to increase the performance of applications in
an application centeric way~\cite{che2009rodinia, tolke2008teraflop, suchard2010understanding, walters2009evaluating,
lin2010accelerating, Khorasanigpu}. For instance, Nukada et al. propose an
efficient 3D FFT that uses shared memory to exchange data efficiently between threads~\cite{fftgpu}.

Other work studies more general approaches to harness the benefits of shared memory, typically by
providing libraries or compile time systems that use shared memory as a scratch-pad to optimize the
memory performance of applications~\cite{Baskaran, YangCompiler, ueng2008cuda, Khangpu, moazeni2009memory,
ji2011using}.
For instance CudaDMA provides a library targeting scientific applications that allows the programmer
to stage data in shared memory and use the data from there~\cite{bauer2011cudadma}. A producer
consumer approach is proposed where some warps only load the data in shared memory (producer)
and others only do the computation (consumer). What we achieve with S-L1 can also be achieved with
CudaDMA, however in CudaDMA the programmer is responsible to use the API manually, set the number of
producer and consumer threads, and assign the proper size of shared memory to different threads.

Yang et al. propose a series of compiler optimizations, including vectorization and data
prefetching, to improve the bandwidth of GPU memory~\cite{YangCompiler}. In particular, they provide
a technique in which uncoalesced memory accesses are transformed to coalesced ones using shared
memory for staging.

Others have also studied the characteristics of GPU memory and GPU
caches~\cite{wongDomestifying, Gaurgpu}. Jia et al.
characterize GPU L1 cache locality in Nvidia GPUs and provide a taxonomy for reasoning about
different types of access patterns and how they might benefit from L1
caches~\cite{jia2012characterizing}.
Tore et al. provides insights into how to tune the configuration of GPU threads to achieve higher
cache hit rates and also offers an observation on how the L1 impacts a handful of simple
kernels~\cite{torres2011understanding}.

Li et al. proposes that register files are a better storage for thread-private
data than shared memory~\cite{ispasswork}. As an experiment to test this, we
modified S-L1 to use registers files as its data storage and observed that its
performance degrades. The reason, we found, is that GPU registers are small
(e.g. 4-bytes) and therefore, a cache-line spans multiple registers, which will
cause the entire cache line (e.g. four registers) to be copied to local memory
before a location of it is accessed, if the accesses are dynamic (i.e. not know
at compile time) -- which is the case in S-L1.


%They also provide a compile-time algorithm to identify an application's memory access pattern and to use
%that information to configure cache usage (by not caching some memory accesses). The disadvantage of
%their approach is that they still use the hardware L1 cache which gets thrashed even when memory
%accesses are to only one data structure. Moreover, since their analysis is done at compile time,
%they are unable to capture any locality with input data dependence.


%Wong et al. uses michrobenchmarking to reverse engineer the undocumented features of NVIDIA GPUs,
%including their memory structures and cache parameters~\cite{wongDomestifying}.

%A study conducted before GPU L1 caches were introduced, proposes a software-managed L1 cache for
%sum-product solver application. The proposed cache is read-only and direct-map, where the entire
%cache space is shared among all threads~\cite{silberstein2008efficient} \todo{we can exclude this}.

Finally, some studied the potential architectural changes that could improve the GPU caching
behavior, including a recent study that analyzes potential coherent models for GPU L1
caches~\cite{singh2013cache}.
